---
title: "A Comparison Between the spca and PMA packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spca-package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 7,
  fig.align = "center",
  out.width = "70%"
)
```

This vignette compares the results from the **PMA** package's implementation of sparse principal components analysis (SPCA) to the implementation in the **spca** package. Both these are based on the $L_1$-penalized matrix decomposition of Witten, Hastie and Tibshirani (2009). The **PMA** package also includes functions for other analysis methods, such as canonical correlation analysis (CCA), and the fused lasso penalty. These are not currently implemented in **spca**.

We will use the following packages:
```{r setup, message = FALSE}
library("spca")
library("PMA")
```


## Application to random data

We will use a random data matrix to compare the two implementations:

```{r}
set.seed(801)
X <- matrix(rnorm(100), nrow = 20, ncol = 5)
```

We column-centre this so that we do not have to ask the PCA implementations to do it first:

```{r}
X <- sweep(X, MARGIN = 2, STATS = colMeans(X), FUN = `-`)
```

This is important since the implementations use different standardizations. See *Details* under `?spca` for some information.

We start by conducting sparse PCA on `X` using the **spca** package:

```{r}
spca_X <- mspca(X, k = 5, c = 1.25, center = FALSE, scale = FALSE)
```

Here the value of $c$ must be in the interval $[1, \sqrt{5}]$, so that 1.25 should result is a relatively sparse results. 

We can do the same analysis using `SPC()` from **PMA**:

```{r}
PMA_X <- SPC(X, K = 5, sumabsv = 1.25, trace = FALSE, center = FALSE, niter = 100)
```

## Comparing the loadings

The loadings for **spca** are:

```{r}
spca_X$loadings
```

The loadings for **PMA** are:

```{r}
PMA_X$v
```

These are similar, but not identical. 

The $L_1$-norms of the columns are as follows:

```{r}
colSums(abs(spca_X$loadings))
colSums(abs(PMA_X$v))
```

The $L_2$-norms of the columns are as follows:
```{r}
colSums(spca_X$loadings^2)
colSums(PMA_X$v^2)
```

Clearly, both solutions satisfy the constraints.

## Comparing the explained variance

The explained variance for **spca** is:

```{r}
spca_X$pve
```

The result for **PMA** is:

```{r}
PMA_X$prop.var.explained
```


## Final remarks


## References

Witten, D. M., Tibshirani, R., & Hastie, T. (2009). A penalized matrix decomposition, with applications to 
sparse principal components and canonical correlation analysis. *Biostatistics*, 10(3), 515-534.
